Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

### 2024-06-16 EBDVF-2024/paper.qmd
edits

---
title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
author:
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    affiliations: 
        - id: swc
          name: The Semantic Web Company
          country: Austria
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliations: 
        - id: onto
          name: Ontotext
          country: Bulgaria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
    affiliations:
        - ref: onto
abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
  The Vocabulary Hub service (part of the IDS RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.
keywords: 
  - dataspaces
  - semantic interoperability
  - semantic technologies
  - ontologies
  - vocabulary hub
  - oil and gas
  - renewable energy
  - refineries
  - windfarms
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Workshop on Semantic Interoperability in Data Spaces 
      formatting: preprint
      model: 3p
      cite-style: super
---

@@

## IDS Vocabulary Hub
In this paper, we focus on the IDSA approach to dataspaces. 
IDSA defines the architecture of dataspaces in the reference architecture model IDS RAM (current version 4). 
One of the defined architectural components, the Vocabulary Hub, has the role of providing vocabularies to annotate and describe data assets and services. 
These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions. 
It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies 
and thereby provides semantic interoperability in a standardized and machine-readable way.

The Vocabulary Hub is one of the core components of any dataspace not only those designed in accordance with IDS RAM. 
Other reference architecture models, such as GAIA-X architecture document 22.10 Release [@GaiaXArchitectureDocument2022] and Fraunhofer's Reference Architecture Model for the Industrial Data Space [@ReferenceArhictectureModel2017] also describe components with a similar role - the Federated Catalogue and Broker Service, respectively. 
Therefore, our approach designed around the IDS RAM is adaptable and applicable to data spaces designed in accordance with other reference architecture models.

## GraphDB Graph Database
Ontotext GraphDB is a highly efficient, scalable and robust graph database with RDF and SPARQL support.
It also offers data virtualization capabilities which enable direct access using SPARQL to relational databases, eliminating the need to replicate data.
It can perform semantic inferencing at scale, which enables deriving new semantic facts from existing facts.

GraphDB handles massive loads, queries, and inferencing in real time.
Reasoning and query evaluation are performed over a persistent storage layer. Loading, reasoning, and query evaluation are executed extremely fast even against huge ontologies and knowledge bases.
According to the [LDBC Semantic Publishing Benchmark](https://ldbcouncil.org/benchmarks/spb/), it is one of the most scalable OWL repositories currently available.

## PoolParty Semantic Suite
PoolParty Semantic Suite is a semantic middleware platform based on W3C standards, specifically the Semantic Web, which provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.

At the core of PoolParty is the Thesaurus Manager, which can be used to create and maintain RDF-based vocabularies, including ontologies, schemas and different kinds of knowledge models. These vocabularies can be applied for various use cases, from bridging data silos in enterprises by implementing data consolidation via the vocabularies, to annotating documents with semantic annotations to implement semantic search and recommendations which provide insights into data.

@@

## The GraphDB Poolparty Integration
The integration of GraphDB and PoolParty leverages synergies of a high-performance graph database with a platform for knowledge graph management.

Both GraphDB and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL. They therefore seamlessly integrate with each other, but also with other services and components in dataspaces which support these machine-readable standards for semantic interoperability.

@@

## Semantic Layer Approach
The Vocabulary Hub, as IDS defines it in the IDS RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
Those vocabularies are to be expressed in RDF and by syntactically compatible with OWL to allow for expressive semantics. 
The IDS RAM does not require, but nonetheless allows for, extending the function of the Vocabulary hub towards providing ontology mappings [@FunctionalLayerIDSRAM2023b], which would enable dataspace connectors to automatically convert data between user-specific and standard data formats. Such automatic conversions are to be performed by Data Tranformation Apps by either the Data Consumers, the Data Providers, or both.

We aim to achieve this through a Semantic Layer for dataspaces, 
which provides vocabulary-based services for advanced and automated semantic descriptions of metadata and data. 
The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.

In the following, we discuss

i) the services provided for the Vocabulary Hub component and 
ii) additional services as Vocabulary Hub extensions that implement a Semantic Layer for dataspaces.

@@

- A Data Provider sends data through the Connector, 
  and at the same time provides basic metadata (DCAT)
- Selected data can be stored in Database(s) managed by the dataspace, 
  for easier access and querying.
  - For large-volume time series data, we recommend to use InfluxDB, PostgreSQL, Google BigQuery, etc.
  - Of course, access control is very important here. 
    We'll leverage GraphDB's [Fine-grained Access Control](https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html), 
    but access control in the time series database is also important.
  - Depending on Provider preferences and agreements, 
    other data may stay at the source, 
    and be delivered to the Consumer only dynamically through the Connector.
- The Vocabulary Hub stores relevant semantic assets: ontologies and thesauri. 
  These are added based on relevant use cases and datasets, following a dataspace governance process.
- The Vocabulary Hub also stores mapping assets, such as OBDA or R2RML mappings and CSVW manifests.
- PoolParty Semantic Integrator is used to edit, view and manage the semantic assets, 
  and to bind mapping assets to incoming data.
- For virtual access to relational data, ONTOP is used to translate SPARQL to SQL dynamically 
  and to convert returned data to SPARQL result format.
- The Data Consumer benefits from:
  - More powerful dataset discovery by using richer semantic queries, e.g.
    - "Give me all time series related to compressor C123" (equipment instance)
    - "Give me all time series related to temperature" (measured quantity)
    - "Give me all time series of bearings" (equipment kind or machine part)
  - Harmonized querying of datasets from multiple providers.

>>>>

---
title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
author:
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    corresponding: true
    affiliations:
      - id: swc
        name: The Semantic Web Company
        country: Austria
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliations:
      - id: onto
        name: Ontotext
        country: Bulgaria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
    corresponding: true
    affiliations:
      - ref: onto
abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
  The Vocabulary Hub service (part of the IDS RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.
keywords:
  - dataspaces
  - semantic interoperability
  - semantic technologies
  - ontologies
  - vocabulary hub
  - oil and gas
  - renewable energy
  - refineries
  - windfarms
journal:
  highlights: # This makes a first page only with the Elsevier highlights: maybe we should remove it
    - The Vocabulary Hub is used in a dataspace to hold semantic vocabularies (ontologies and thesauri).
    - Proposes and implements a Semantic Layers architecture that allows binding semantic vocabularies to incoming datasets.
    - Uses semantic mapping languages (OBDA, R2RML, CSVW) to harmonize datasets dynamically.
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Workshop on Semantic Interoperability in Data Spaces
      formatting: preprint
      model: 3p
      cite-style: super
---

@@

## IDS Vocabulary Hub
IDSA defines the dataspace architecture in the IDS RAM [@IDSRAM2024].
The Vocabulary Hub is one of the defined architectural components that provides vocabularies to annotate and describe data assets and services.
These vocabularies form a common language to enable semantic interoperability among dataspace participants.
The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions.
It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies
and thereby provides semantic interoperability in a standardized and machine-readable way.

Other architecture reference models, such as the GAIA-X Architecture [@GaiaXArchitectureDocument2022] 
describe components with a similar role: the Federated Catalogue and Broker Service.
Therefore our approach  is adaptable and applicable to dataspaces designed in accordance with other reference architecture models as well.

## GraphDB Graph Database
[Ontotext GraphDB](https://graphdb.ontotext.com/) is a highly efficient, scalable and robust graph database with RDF and SPARQL support.
It also offers data virtualization capabilities that enable direct SPARQL access to relational databases, eliminating the need to replicate data.
It can perform semantic reasoning at scale and supports both incremental inference when triples are inserted, and retraction when triples are deleted.
GraphDB handles massive semantic data loads, queries, and inferencing.
Reasoning and query evaluation are performed over a persistent storage layer. 
According to the [LDBC Semantic Publishing Benchmark](https://ldbcouncil.org/benchmarks/spb/), GraphDB is one of the most scalable OWL repositories currently available.

In addition to numerous commercial projects, 
GraphDB has been used for a variety of research projects, from astronomy to zoology [@alexievDiverseUsesSemantic2021].

## PoolParty Semantic Suite
[PoolParty Semantic Suite](PoolParty Semantic Suite) is a semantic middleware platform based on W3C Semantic Web standards 
that provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.

At the core of PoolParty is the Thesaurus Manager that is used to create and maintain RDF-based vocabularies (thesauri and ontologies), 
schemas and different kinds of knowledge models. 
These vocabularies can be applied for various use cases, 
from bridging data silos in enterprises by implementing data consolidation via the vocabularies, 
to annotating documents with semantic annotations to implement semantic search and recommendations which provide insights into data.

@@

## The GraphDB Poolparty Integration
The integration of GraphDB and PoolParty leverages synergies of a high-performance graph database with a knowledge graph management platform.

Both GraphDB and PoolParty are based on W3C recommendations (RDF, RDFS, OWL, SPARQL etc). 
They therefore seamlessly integrate with each other, but also with other services and components in dataspaces which support these machine-readable standards for semantic interoperability.

@@

## Semantic Layer Approach

The Vocabulary Hub is defined in the IDS RAM as a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies.
Vocabularies are expressed in RDF and use RDFS and OWL for ontologies, and SKOS for thesauri.
The IDS RAM does not require, but nonetheless allows extending the function of the Vocabulary Hub towards providing ontology mappings [@IDSARAMFunctional2023]
that enable dataspace connectors to automatically convert data between user-specific and standard data formats.
Such automatic conversions can be performed by Data Tranformation Apps by the Data Consumer, the Data Provider, or both.

We aim to achieve this through a Semantic Layer for dataspaces
that provides vocabulary-based services for advanced and automated semantic description of metadata and data.
The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.

In the following, we discuss
i) the services provided for the Vocabulary Hub component and
ii) additional services as Vocabulary Hub extensions that implement a Semantic Layer for dataspaces.

@@

- A Data Provider sends data through the Connector,
  and at the same time provides basic metadata (DCAT)
- Selected data can be stored in Database(s) managed by the dataspace,
  for easier access and querying.
  - For large-volume time series data, we recommend to use InfluxDB, PostgreSQL, Google BigQuery, etc.
  - Of course, access control is very important here.
    We'll leverage GraphDB's [Fine-grained Access Control](https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html),
    but access control in the time series database is also important.
  - Depending on Provider preferences and agreements,
    other data may stay at the source,
    and be delivered to the Consumer only dynamically through the Connector.
- The Vocabulary Hub stores relevant semantic assets: ontologies and thesauri.
  These are added based on relevant use cases and datasets, following a dataspace governance process.
- The Vocabulary Hub also stores mapping assets, such as OBDA or R2RML mappings and CSVW manifests.
- PoolParty Semantic Integrator is used to edit, view and manage the semantic assets,
  and to bind mapping assets to incoming data.
- For virtual access to relational data, ONTOP is used to translate SPARQL to SQL dynamically
  and to convert returned data to SPARQL result format.

The Data Consumer benefits from:
- More powerful dataset discovery by using richer semantic queries, e.g.
  - "Give me all time series related to compressor C123" (equipment instance)
  - "Give me all time series related to temperature" (measured quantity)
  - "Give me all time series of bearings" (equipment kind or machine part)
- Harmonized querying of datasets from multiple providers.

### 2024-06-16 EBDVF-2024/paper.tex
edits

\begin{frontmatter}
\title{Raising the Role of Vocabulary Hubs for Semantic Data
Interoperability in Dataspaces}
\author[1]{Robert David%
%
}
 \ead{robert.david@semantic-web.com} 
\author[2]{Petar Ivanov%
%
}
 \ead{petar.ivanov@ontotext.com} 
\author[2]{Vladimir Alexiev%
%
}
 \ead{vladimir.alexiev@ontotext.com}

@@

@@

In this paper, we focus on the IDSA approach to dataspaces. IDSA defines
the architecture of dataspaces in the reference architecture model IDS
RAM (current version 4). One of the defined architectural components,
the Vocabulary Hub, has the role of providing vocabularies to annotate
and describe data assets and services. These vocabularies form a common
language to enable semantic interoperability among the dataspace
participants. The Vocabulary Hub is responsible for resolving
annotations of data assets and for providing details in the form of
standardized semantic descriptions. It supports the use of Semantic Web
standards for defining and representing RDF vocabularies and ontologies
and thereby provides semantic interoperability in a standardized and
machine-readable way.

The Vocabulary Hub is one of the core components of any dataspace not
only those designed in accordance with IDS RAM. Other reference
architecture models, such as GAIA-X architecture document 22.10 Release
\citep{GaiaXArchitectureDocument2022} and Fraunhofer's Reference
Architecture Model for the Industrial Data Space
\citep{ReferenceArhictectureModel2017} also describe components with a
similar role - the Federated Catalogue and Broker Service, respectively.
Therefore, our approach designed around the IDS RAM is adaptable and
applicable to data spaces designed in accordance with other reference
architecture models.

@@

Ontotext GraphDB is a highly efficient, scalable and robust graph
database with RDF and SPARQL support. It also offers data virtualization
capabilities which enable direct access using SPARQL to relational
databases, eliminating the need to replicate data. It can perform
semantic inferencing at scale, which enables deriving new semantic facts
from existing facts.

GraphDB handles massive loads, queries, and inferencing in real time.
Reasoning and query evaluation are performed over a persistent storage
layer. Loading, reasoning, and query evaluation are executed extremely
fast even against huge ontologies and knowledge bases. According to the
\href{https://ldbcouncil.org/benchmarks/spb/}{LDBC Semantic Publishing
Benchmark}, it is one of the most scalable OWL repositories currently
available.

@@

PoolParty Semantic Suite is a semantic middleware platform based on W3C
standards, specifically the Semantic Web, which provides a wide variety
of functionality in the area of knowledge graph management, graph-based
NLP, semantic search and recommender systems.

At the core of PoolParty is the Thesaurus Manager, which can be used to
create and maintain RDF-based vocabularies, including ontologies,
schemas and different kinds of knowledge models. These vocabularies can
be applied for various use cases, from bridging data silos in
enterprises by implementing data consolidation via the vocabularies, to
annotating documents with semantic annotations to implement semantic
search and recommendations which provide insights into data.

@@

The integration of GraphDB and PoolParty leverages synergies of a
high-performance graph database with a platform for knowledge graph
management.

Both GraphDB and PoolParty are based on W3C recommendations like RDF,
OWL and SPARQL. They therefore seamlessly integrate with each other, but
also with other services and components in dataspaces which support
these machine-readable standards for semantic interoperability.

@@

The Vocabulary Hub, as IDS defines it in the IDS RAM, is a basic
building block to achieve semantic interoperability in dataspaces by
provisioning common vocabularies. Those vocabularies are to be expressed
in RDF and by syntactically compatible with OWL to allow for expressive
semantics. The IDS RAM does not require, but nonetheless allows for,
extending the function of the Vocabulary hub towards providing ontology
mappings \citep{FunctionalLayerIDSRAM2023b}, which would enable
dataspace connectors to automatically convert data between user-specific
and standard data formats. Such automatic conversions are to be
performed by Data Tranformation Apps by either the Data Consumers, the
Data Providers, or both.

We aim to achieve this through a Semantic Layer for dataspaces, which
provides vocabulary-based services for advanced and automated semantic
descriptions of metadata and data. The Semantic Layer is implemented as
services based on the GraphDB and PoolParty and provided as a one
seamlessly integrated component.

In the following, we discuss

\begin{enumerate}
\def\labelenumi{\roman{enumi})}
\tightlist
\item
  the services provided for the Vocabulary Hub component and
\item
  additional services as Vocabulary Hub extensions that implement a
  Semantic Layer for dataspaces.
\end{enumerate}

@@

  \begin{itemize}
  \tightlist
  \item
    For large-volume time series data, we recommend to use InfluxDB,
    PostgreSQL, Google BigQuery, etc.
  \item
    Of course, access control is very important here. We'll leverage
    GraphDB's
    \href{https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html}{Fine-grained
    Access Control}, but access control in the time series database is
    also important.
  \item
    Depending on Provider preferences and agreements, other data may
    stay at the source, and be delivered to the Consumer only
    dynamically through the Connector.
  \end{itemize}
\item
  The Vocabulary Hub stores relevant semantic assets: ontologies and
  thesauri. These are added based on relevant use cases and datasets,
  following a dataspace governance process.
\item
  The Vocabulary Hub also stores mapping assets, such as OBDA or R2RML
  mappings and CSVW manifests.
\item
  PoolParty Semantic Integrator is used to edit, view and manage the
  semantic assets, and to bind mapping assets to incoming data.
\item
  For virtual access to relational data, ONTOP is used to translate
  SPARQL to SQL dynamically and to convert returned data to SPARQL
  result format.
\item
  The Data Consumer benefits from:

  \begin{itemize}
  \tightlist
  \item
    More powerful dataset discovery by using richer semantic queries,
    e.g.

    \begin{itemize}
    \tightlist
    \item
      ``Give me all time series related to compressor C123'' (equipment
      instance)
    \item
      ``Give me all time series related to temperature'' (measured
      quantity)
    \item
      ``Give me all time series of bearings'' (equipment kind or machine
      part)
    \end{itemize}
  \item
    Harmonized querying of datasets from multiple providers.
  \end{itemize}
\end{itemize}

>>>>

\begin{frontmatter}
\title{Raising the Role of Vocabulary Hubs for Semantic Data
Interoperability in Dataspaces}
\author[1]{Robert David%
\corref{cor1}%
}
 \ead{robert.david@semantic-web.com} 
\author[2]{Petar Ivanov%
%
}
 \ead{petar.ivanov@ontotext.com} 
\author[2]{Vladimir Alexiev%
\corref{cor1}%
}
 \ead{vladimir.alexiev@ontotext.com}

@@

\begin{highlights}
\item The Vocabulary Hub is used in a dataspace to hold semantic
vocabularies (ontologies and thesauri).\item Proposes and implements a
Semantic Layers architecture that allows binding semantic vocabularies
to incoming datasets.\item Uses semantic mapping languages (OBDA, R2RML,
CSVW) to harmonize datasets dynamically.
\end{highlights}

@@

IDSA defines the dataspace architecture in the IDS RAM
\citep{IDSRAM2024}. The Vocabulary Hub is one of the defined
architectural components that provides vocabularies to annotate and
describe data assets and services. These vocabularies form a common
language to enable semantic interoperability among dataspace
participants. The Vocabulary Hub is responsible for resolving
annotations of data assets and for providing details in the form of
standardized semantic descriptions. It supports the use of Semantic Web
standards for defining and representing RDF vocabularies and ontologies
and thereby provides semantic interoperability in a standardized and
machine-readable way.

Other architecture reference models, such as the GAIA-X Architecture
\citep{GaiaXArchitectureDocument2022} describe components with a similar
role: the Federated Catalogue and Broker Service. Therefore our approach
is adaptable and applicable to dataspaces designed in accordance with
other reference architecture models as well.

@@

\href{https://graphdb.ontotext.com/}{Ontotext GraphDB} is a highly
efficient, scalable and robust graph database with RDF and SPARQL
support. It also offers data virtualization capabilities that enable
direct SPARQL access to relational databases, eliminating the need to
replicate data. It can perform semantic reasoning at scale and supports
both incremental inference when triples are inserted, and retraction
when triples are deleted. GraphDB handles massive semantic data loads,
queries, and inferencing. Reasoning and query evaluation are performed
over a persistent storage layer. According to the
\href{https://ldbcouncil.org/benchmarks/spb/}{LDBC Semantic Publishing
Benchmark}, GraphDB is one of the most scalable OWL repositories
currently available.

In addition to numerous commercial projects, GraphDB has been used for a
variety of research projects, from astronomy to zoology
\citep{alexievDiverseUsesSemantic2021}.

@@

\href{PoolParty\%20Semantic\%20Suite}{PoolParty Semantic Suite} is a
semantic middleware platform based on W3C Semantic Web standards that
provides a wide variety of functionality in the area of knowledge graph
management, graph-based NLP, semantic search and recommender systems.

At the core of PoolParty is the Thesaurus Manager that is used to create
and maintain RDF-based vocabularies (thesauri and ontologies), schemas
and different kinds of knowledge models. These vocabularies can be
applied for various use cases, from bridging data silos in enterprises
by implementing data consolidation via the vocabularies, to annotating
documents with semantic annotations to implement semantic search and
recommendations which provide insights into data.

@@

The integration of GraphDB and PoolParty leverages synergies of a
high-performance graph database with a knowledge graph management
platform.

Both GraphDB and PoolParty are based on W3C recommendations (RDF, RDFS,
OWL, SPARQL etc). They therefore seamlessly integrate with each other,
but also with other services and components in dataspaces which support
these machine-readable standards for semantic interoperability.

@@

The Vocabulary Hub is defined in the IDS RAM as a basic building block
to achieve semantic interoperability in dataspaces by provisioning
common vocabularies. Vocabularies are expressed in RDF and use RDFS and
OWL for ontologies, and SKOS for thesauri. The IDS RAM does not require,
but nonetheless allows extending the function of the Vocabulary Hub
towards providing ontology mappings \citep{IDSARAMFunctional2023} that
enable dataspace connectors to automatically convert data between
user-specific and standard data formats. Such automatic conversions can
be performed by Data Tranformation Apps by the Data Consumer, the Data
Provider, or both.

We aim to achieve this through a Semantic Layer for dataspaces that
provides vocabulary-based services for advanced and automated semantic
description of metadata and data. The Semantic Layer is implemented as
services based on the GraphDB and PoolParty and provided as a one
seamlessly integrated component.

In the following, we discuss i) the services provided for the Vocabulary
Hub component and ii) additional services as Vocabulary Hub extensions
that implement a Semantic Layer for dataspaces.

@@

  \begin{itemize}
  \tightlist
  \item
    For large-volume time series data, we recommend to use InfluxDB,
    PostgreSQL, Google BigQuery, etc.
  \item
    Of course, access control is very important here. We'll leverage
    GraphDB's
    \href{https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html}{Fine-grained
    Access Control}, but access control in the time series database is
    also important.
  \item
    Depending on Provider preferences and agreements, other data may
    stay at the source, and be delivered to the Consumer only
    dynamically through the Connector.
  \end{itemize}
\item
  The Vocabulary Hub stores relevant semantic assets: ontologies and
  thesauri. These are added based on relevant use cases and datasets,
  following a dataspace governance process.
\item
  The Vocabulary Hub also stores mapping assets, such as OBDA or R2RML
  mappings and CSVW manifests.
\item
  PoolParty Semantic Integrator is used to edit, view and manage the
  semantic assets, and to bind mapping assets to incoming data.
\item
  For virtual access to relational data, ONTOP is used to translate
  SPARQL to SQL dynamically and to convert returned data to SPARQL
  result format.
\end{itemize}

The Data Consumer benefits from: - More powerful dataset discovery by
using richer semantic queries, e.g. - ``Give me all time series related
to compressor C123'' (equipment instance) - ``Give me all time series
related to temperature'' (measured quantity) - ``Give me all time series
of bearings'' (equipment kind or machine part) - Harmonized querying of
datasets from multiple providers.
